---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

<!-- badges: start -->
[![R-CMD-check](https://github.com/massimoaria/e2tree/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/massimoaria/e2tree/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  dpi = 300
)
```

## Setup

You can install the developer version of e2tree from [GitHub](https://github.com) with:

```{r eval=FALSE}
install.packages("remotes")
remotes::install_github("massimoaria/e2tree")
```

```{r warning=FALSE, message=FALSE}
require(e2tree)
require(tidyverse)
options(dplyr.summarise.inform = FALSE)
require(randomForest)
require(Matrix)
require(future.apply)
```

```{r set-theme, include=FALSE}
theme_set(
  theme_classic() +
    theme(
      plot.background = element_rect(fill = "transparent", colour = NA),
      panel.background = element_rect(fill = "transparent", colour = NA)
    )
)
knitr::opts_chunk$set(dev.args = list(bg = "transparent"))
```

## Warnings

The package is still under development and therefore, for the time being, there are the following limitations:

- Only ensembles trained with the randomForest package are supported. Additional packages and approaches will be supported in the future;

- Currently e2tree works only in the case of classification problems. It will gradually be extended to other problems related to the nature of the response variable: regression, counting, multivariate response, etc.


## Example 1: IRIS dataset

In this example, we want to show the main functions of the e2tree package.

Starting from the IRIS dataset, we will train an ensemble tree using the randomForest package and then subsequently use e2tree to obtain an explainable tree synthesis of the ensemble classifier.

```{r}
# Set random seed to make results reproducible:
set.seed(0)

# Initialize the split
iris_split <- iris %>% initial_split(prop = 0.6)
iris_split
# Assign the data to the correct sets
training <- iris_split %>% training()
validation <- iris_split %>% testing()
response_training <- training[,5]
response_validation <- validation[,5]

```


Train an Random Forest model with 1000 weak learners

```{r}
# Perform training:
ensemble = randomForest(Species ~ ., data = training, importance = TRUE, proximity = TRUE)
```

Here, we create the dissimilarity matrix between observations through the createDisMatrix function

```{r}
D = createDisMatrix(ensemble, data = training, label = "Species", parallel = TRUE)
#dis <- 1-rf$proximity
```

setting e2tree parameters

```{r}
setting=list(impTotal=0.1, maxDec=0.01, n=5, level=5, tMax=5)
```


Build an explainable tree for RF

```{r}
tree <- e2tree(Species ~ ., data = training, D, ensemble, setting)

```


Let's have a look at the output

```{r}
tree %>% glimpse()
```

Prediction with the new tree (example on training)

```{r}
pred <- ePredTree(tree, training[,-5], target="virginica")
```

Comparison of predictions (training sample) of RF and e2tree

```{r}
table(pred$fit, ensemble$predicted)
```

Comparison of predictions (training sample) of RF and correct response

```{r}
table(ensemble$predicted, response_training)
```

Comparison of predictions (training sample) of e2tree and correct response

```{r}
table(pred$fit,response_training)
```

Variable Importance

```{r}
ensemble_imp <- ensemble$importance %>% as.data.frame %>% 
  mutate(Variable = rownames(ensemble$importance),
         RF_Var_Imp = round(MeanDecreaseAccuracy,2)) %>% 
  select(Variable, RF_Var_Imp)

V <- vimp(tree, training)
#V <- V$vimp %>% 
#  select(Variable,MeanImpurityDecrease, `ImpDec_ setosa`, `ImpDec_ versicolor`,`ImpDec_ virginica`) %>% 
#  mutate_at(c("MeanImpurityDecrease","ImpDec_ setosa", "ImpDec_ versicolor","ImpDec_ virginica"), round,2) %>% 
#  left_join(ensemble_imp, by = "Variable") %>% 
#  select(Variable, RF_Var_Imp, MeanImpurityDecrease, starts_with("ImpDec")) %>% 
#  rename(ETree_Var_Imp = MeanImpurityDecrease)

V
  
```


Comparison with the validation sample

```{r}
ensemble.pred <- predict(ensemble, validation[,-5], proximity = TRUE)

pred_val<- ePredTree(tree, validation[,-5], target="virginica")
```

Comparison of predictions (sample validation) of RF and e2tree

```{r}
table(pred_val$fit, ensemble.pred$predicted)
```

Comparison of predictions (validation sample) of RF and correct response

```{r}
table(ensemble.pred$predicted, response_validation)
ensemble.prob <- predict(ensemble, validation[,-5], proximity = TRUE, type="prob")
roc_ensemble<- roc(response_validation, ensemble.prob$predicted[,"virginica"], target="virginica")
roc_ensemble$auc
```

Comparison of predictions (validation sample) of e2tree and correct response

```{r}
table(pred_val$fit, response_validation)
roc_res <- roc(response_validation, pred_val$score, target="virginica")
roc_res$auc
```

