---
title: "Introduction to e2tree: Explainable Ensemble Trees"
author: "Massimo Aria, Agostino Gnasso"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to e2tree: Explainable Ensemble Trees}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

The **e2tree** package implements the Explainable Ensemble Tree (E2Tree)
methodology, which constructs a single interpretable decision tree that
approximates the predictions of an ensemble model (such as a random forest).
The key idea is to use the dissimilarity structure induced by the ensemble
to guide the construction of the tree, thereby preserving the predictive
information encoded in the ensemble while providing a transparent,
human-readable model.

The methodology is described in:

- Aria, M., Gnasso, A., Iorio, C., & Pandolfo, G. (2024). Explainable
  ensemble trees. *Computational Statistics*, 39(1), 3--19.
- Aria, M., Gnasso, A., Iorio, C., & Fokkema, M. (2026). Extending
  Explainable Ensemble Trees to Regression Contexts. *Applied Stochastic
  Models in Business and Industry*, 42(1), e70064.

## Installation

```{r eval = FALSE}
# Install from CRAN (when available):
install.packages("e2tree")

# Or install the development version from GitHub:
# devtools::install_github("massimoaria/e2tree")
```

## Workflow

A typical e2tree analysis follows these steps:

1. **Train an ensemble model** (random forest via `randomForest` or `ranger`)
2. **Compute the dissimilarity matrix** from the ensemble using `createDisMatrix()`
3. **Build the E2Tree** using `e2tree()`
4. **Inspect and visualize** the tree (print, summary, plot)
5. **Predict** on new data with `predict()`
6. **Validate** the E2Tree structure with `eValidation()` and `loi()`

## Classification Example

We use the classic `iris` dataset to demonstrate a classification workflow.

### Step 1: Prepare data and train ensemble

```{r classification-setup}
library(e2tree)

data(iris)
set.seed(42)
smp_size <- floor(0.75 * nrow(iris))
train_ind <- sample(seq_len(nrow(iris)), size = smp_size)
training <- iris[train_ind, ]
validation <- iris[-train_ind, ]
```

```{r classification-ensemble}
library(randomForest)
ensemble <- randomForest(Species ~ ., data = training,
                         importance = TRUE, proximity = TRUE)
ensemble
```

### Step 2: Compute dissimilarity matrix

The ensemble induces a co-occurrence structure that captures how frequently
pairs of observations fall into the same terminal nodes across the ensemble's
trees. This structure is encoded in the co-occurrence matrix
$\mathbf{O} = [o_{ij}]$, where each entry $o_{ij} \in [0, 1]$ reflects the
weighted frequency with which observations $i$ and $j$ are grouped together.
The corresponding dissimilarity matrix is $\mathbf{D} = \mathbf{1} - \mathbf{O}$.

```{r classification-dismatrix}
D <- createDisMatrix(ensemble, data = training, label = "Species",
                     parallel = list(active = FALSE, no_cores = 1))
dim(D)
```

### Step 3: Build the E2Tree

The `e2tree()` function constructs a single tree that best represents the
ensemble's dissimilarity structure. The `setting` argument controls the
stopping rules:

- `impTotal`: minimum impurity to attempt a split
- `maxDec`: minimum decrease in impurity required
- `n`: minimum number of observations in a node
- `level`: maximum tree depth

```{r classification-tree}
setting <- list(impTotal = 0.1, maxDec = 0.01, n = 2, level = 5)
tree <- e2tree(Species ~ ., training, D, ensemble, setting)
```

### Step 4: Inspect the tree

```{r classification-print}
print(tree)
```

```{r classification-summary}
summary(tree)
```

The `nodes()` accessor extracts the tree structure as a data frame:

```{r classification-nodes}
# Terminal nodes only
nodes(tree, terminal = TRUE)
```

### Step 5: Visualization

The `plot()` method renders the tree using `rpart.plot`:

```{r classification-plot, fig.width = 8, fig.height = 6}
plot(tree, ensemble = ensemble, main = "E2Tree - Iris Classification")
```

You can also convert the tree to `rpart` or `partykit` formats for
alternative visualizations:

```{r classification-rpart, fig.width = 8, fig.height = 6}
# Convert to rpart
rpart_obj <- as.rpart(tree, ensemble)

# Convert to partykit (if installed)
if (requireNamespace("partykit", quietly = TRUE)) {
  party_obj <- partykit::as.party(tree)
  plot(party_obj)
}
```

### Step 6: Prediction

Use the standard `predict()` method on new data:

```{r classification-predict}
pred <- predict(tree, newdata = validation)
head(pred)
```

Fitted values for the training data:

```{r classification-fitted}
head(fitted(tree))
```

### Step 7: Variable importance

```{r classification-vimp}
vi <- vimp(tree, data = training)
vi$vimp
vi$g_imp
```

## Regression Example

The same workflow applies to regression problems. Here we use the `mtcars` dataset.

```{r regression-setup}
data(mtcars)
set.seed(123)
smp_size <- floor(0.75 * nrow(mtcars))
train_ind <- sample(seq_len(nrow(mtcars)), size = smp_size)
training_reg <- mtcars[train_ind, ]
validation_reg <- mtcars[-train_ind, ]
```

```{r regression-ensemble}
ensemble_reg <- randomForest(mpg ~ ., data = training_reg, ntree = 500,
                             importance = TRUE, proximity = TRUE)
```

```{r regression-dismatrix}
D_reg <- createDisMatrix(ensemble_reg, data = training_reg, label = "mpg",
                         parallel = list(active = FALSE, no_cores = 1))
```

```{r regression-tree}
setting_reg <- list(impTotal = 0.1, maxDec = 1e-6, n = 2, level = 5)
tree_reg <- e2tree(mpg ~ ., training_reg, D_reg, ensemble_reg, setting_reg)
print(tree_reg)
```

```{r regression-plot, fig.width = 8, fig.height = 6}
plot(tree_reg, ensemble = ensemble_reg, main = "E2Tree - MPG Regression")
```

For regression models, predictions include the standard deviation within
each terminal node, providing a measure of prediction uncertainty:

```{r regression-predict}
pred_reg <- predict(tree_reg, newdata = validation_reg)
head(pred_reg)
```

Residuals are also available:

```{r regression-residuals}
res <- residuals(tree_reg)
summary(res)
```


## Validation of the E2Tree Structure

A fundamental question arises when using E2Tree: how well does the single
interpretable tree capture the original ensemble's structure? The E2Tree
reconstructs the ensemble's co-occurrence matrix $\mathbf{O}$ through
its terminal node structure, producing an approximation
$\hat{\mathbf{O}}_T$. Assessing the quality of this reconstruction is
essential to ensure that the explanatory model faithfully represents
the ensemble's decision logic.

The `eValidation()` function provides two complementary approaches to
validation, selectable through the `test` argument:

- **`test = "mantel"`**: The Mantel test, a classical non-parametric
  test of *association* between similarity matrices.
- **`test = "measures"`**: A family of divergence and similarity measures
  that assess *agreement* between the matrices, with permutation-based
  inference.
- **`test = "both"`** (default): Performs both.

### Association vs. Agreement

The distinction between these two approaches mirrors the classical
distinction between *correlation* and *concordance* in method comparison
studies (Bland & Altman, 1986; Lin, 1989). Two proximity matrices can be
perfectly correlated ($r = 1$) and yet systematically disagree: if
$\hat{\mathbf{O}}_T = c \cdot \mathbf{O}$ for any constant $c \neq 1$,
the Mantel test would declare perfect association, yet the actual
proximity values would not be faithfully reproduced.

This distinction matters in the E2Tree context. The ensemble matrix
$\mathbf{O}$ contains continuous values in $(0, 1)$, reflecting varying
degrees of co-occurrence across the ensemble's trees. The E2Tree matrix
$\hat{\mathbf{O}}_T$, by contrast, has a crisp block-diagonal structure:
observations within the same terminal node share a common co-occurrence
value, while observations in different nodes have zero co-occurrence.
Even when the block structure perfectly matches the grouping pattern of
$\mathbf{O}$, the element-wise values will differ because $\mathbf{O}$
is fuzzy and $\hat{\mathbf{O}}_T$ is crisp. A scale-sensitive divergence
measure can quantify this discrepancy; a correlation-based measure cannot.

### The Mantel Test

The Mantel test evaluates whether the E2Tree proximity matrix is
*associated* with the ensemble proximity matrix. It computes a
correlation-like statistic between the two matrices and assesses
significance through permutation.

```{r val-mantel}
val_mantel <- eValidation(training, tree, D, test = "mantel", graph = FALSE)
print(val_mantel)
```

A significant Mantel test confirms that E2Tree preserves the overall
*pattern* of the ensemble's proximity structure. However, it does not
tell us how closely the actual proximity values are reproduced, nor
does it indicate *where* the reconstruction fails.

### Divergence and Similarity Measures

To assess *agreement* rather than mere association, we use a family of
five measures, each capturing a distinct aspect of reconstruction quality:

| Measure | Type | Range | What it captures |
|---------|------|-------|------------------|
| **nLoI** (normalized Loss of Interpretability) | divergence | [0, 1] | Scale-sensitive per-pair discrepancy with diagnostic decomposition |
| **Hellinger** distance | divergence | [0, 1] | Variance-stabilized element-wise difference (a true metric) |
| **wRMSE** (weighted Root Mean Squared Error) | divergence | [0, 1] | Weighted average discrepancy emphasizing high-proximity regions |
| **RV** coefficient | similarity | [0, 1] | Global structural similarity (matrix-level correlation) |
| **SSIM** (Structural Similarity Index) | similarity | [-1, 1] | Local spatial pattern similarity (luminance, contrast, structure) |

For *divergence* measures, lower values indicate better reconstruction.
For *similarity* measures, higher values indicate better reconstruction.

Statistical significance is assessed through a unified permutation testing
framework based on simultaneous row/column permutation of $\hat{\mathbf{O}}_T$.
This preserves the block-diagonal structure of the E2Tree matrix while
breaking the correspondence between observation labels and node assignments,
providing the appropriate null hypothesis: under $H_0$, E2Tree groups
observations randomly with respect to the ensemble structure.

```{r val-measures}
val_meas <- eValidation(training, tree, D, test = "measures",
                        graph = FALSE, n_perm = 499)
print(val_meas)
```

All divergence measures are significantly lower than their null
expectations (negative Z-statistics), while all similarity measures are
significantly higher (positive Z-statistics), confirming that the E2Tree
faithfully reconstructs the ensemble's proximity structure.

```{r val-measures-table}
measures(val_meas)
```

### Visualizing the Reconstruction

The `plot()` method for `eValidation` objects provides a visual summary
including heatmaps of both proximity matrices, the null distribution of
nLoI (when permutation tests are performed), and the LoI decomposition:

```{r val-plot, fig.width = 8, fig.height = 7}
val_full <- eValidation(training, tree, D, test = "both",
                        graph = FALSE, n_perm = 499)
plot(val_full)
```

### The Normalized Loss of Interpretability (nLoI)

Among the divergence measures, the nLoI deserves special attention because
of its unique *decomposability* into within-node and between-node components.

The nLoI is defined as:
$$\text{nLoI}(\mathbf{O}, \hat{\mathbf{O}}_T) = \frac{1}{M} \sum_{i<j} \frac{(o_{ij} - \hat{o}_{ij})^2}{\max(o_{ij}, \hat{o}_{ij})}$$
where $M = n(n-1)/2$ is the number of unique pairs.

This statistic is connected to the Cressie-Read power divergence family
at $\lambda = -2$ (Neyman-type statistic), with a robustified denominator
that avoids singularities when $\hat{o}_{ij} = 0$ (which is common in
E2Tree due to the block-diagonal structure of $\hat{\mathbf{O}}_T$).

The `loi()` function computes the nLoI and its decomposition directly
from two proximity matrices:

```{r loi-classification}
prox <- proximity(val_full)
result <- loi(prox$ensemble, prox$e2tree)
summary(result)
```

#### LoI Decomposition: Diagnosing Reconstruction Quality

The LoI decomposes into two interpretable components:

- **Within-node component** ($\text{LoI}_{in}$): Measures reconstruction
  fidelity for pairs that E2Tree groups together. It captures the
  calibration error --- whether the single tree assigns co-occurrence
  values that match the ensemble's.

- **Between-node component** ($\text{LoI}_{out}$): Measures the cost of
  the partition. It accumulates the ensemble co-occurrence values
  ($o_{ij}$) for pairs that E2Tree separates into different terminal
  nodes (where $\hat{o}_{ij} = 0$), representing irrecoverable loss.

Since between-node pairs vastly outnumber within-node pairs (especially
with many terminal nodes), the raw totals are not directly comparable.
The per-pair averages (`mean_in` and `mean_out`) provide the meaningful
diagnostic:

- A **high `mean_out`** (e.g., > 0.3) indicates that E2Tree is separating
  pairs with substantial ensemble proximity. The practitioner should
  consider increasing the tree's complexity (more terminal nodes) or
  relaxing pruning constraints.

- A **high `mean_in`** (e.g., > 0.1) indicates poor within-node calibration:
  the partition boundaries are well-placed, but the within-node proximity
  values deviate from the ensemble's. This typically reflects the inherent
  fuzzy-to-crisp structural transition.

- **Low values of both** confirm that the E2Tree faithfully captures the
  ensemble structure --- the partition is well-placed and the calibration
  is accurate.

This diagnostic capability is fundamentally impossible with the Mantel test
or any other scalar association measure, which can only report that the
overall reconstruction is "good" or "poor" without indicating the source
of the discrepancy.

```{r loi-plot}
plot(result)
```

#### Permutation Test for LoI

The `loi_perm()` function provides a standalone permutation test for the
nLoI, useful when a quick assessment of reconstruction significance is
needed without computing all five measures:

```{r loi-perm}
perm <- loi_perm(prox$ensemble, prox$e2tree, n_perm = 499, seed = 42)
print(perm)
plot(perm)
```

### Validation of the Regression Example

The same validation framework applies to regression E2Trees:

```{r val-regression}
val_reg <- eValidation(training_reg, tree_reg, D_reg, test = "measures",
                       graph = FALSE, n_perm = 499)
print(val_reg)
```

Regression E2Trees typically show higher nLoI and Hellinger values than
classification, consistent with the greater complexity of regression
proximity structures and the typically smaller sample sizes. Nonetheless,
all measures should achieve significant p-values, confirming that the
reconstruction captures genuine structure.


## Comparison with Related Packages

The **e2tree** package complements existing R infrastructure for tree-based
models:

- **partykit**: Provides generic classes for representing trees. E2Tree
  objects can be converted to `party` objects via `as.party()`, enabling
  the use of partykit's rich visualization and prediction infrastructure.
- **stablelearner**: Explores the stability of ensemble trees via
  `stabletree()`. While stablelearner focuses on variable and cutpoint
  stability, e2tree constructs a single tree that captures the ensemble's
  dissimilarity-based structure --- a complementary perspective.
- **rpart**: The standard recursive partitioning package. E2Tree objects can
  be converted to `rpart` objects via `as.rpart()` for visualization with
  `rpart.plot`.

## Session Info

```{r session-info}
sessionInfo()
```
